---
title: "SQLite"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(DBI)
library(RSQLite)
library(dbplot)
ontime_dir <- "~/data/ontime"
flight_dir <- "~/data/ontime/flight"
path <- "~/data/ontime"    # '.' is the current folder
db <- "ontime.sqlite"
path_db <- file.path(path, db)
```


## Resources

- [R and Database documentation](https://db.rstudio.com/)
- [Use dplyr with databases](https://db.rstudio.com/r-packages/dplyr/)
- [Visualisation and databases](https://db.rstudio.com/best-practices/visualization/)
- [RSQLite](https://github.com/r-dbi/RSQLite)
- [SQLite browser](https://sqlitebrowser.org/dl/)
- [dbplot vignette](https://rstudio-pubs-static.s3.amazonaws.com/314409_a8af4ecde2a34321a9b6a3a75143f5c8.html)

## SQLite


```{r}
ontime_files <- list.files(ontime_dir, pattern = ".csv", full.names = TRUE)
```


### Delete the database file

```{r}

delete_db_if_exists <- function(path, db) {
  db_exists <- list.files(path, pattern = db)
  if(length(db_exists) != 0) system(paste("rm", path_db))
}
# delete_db_if_exists(path, db)

# it create the db if it does not exists
con <- dbConnect(RSQLite::SQLite(), path_db)

# Encoding of the csv
encoding <- "us-ascii"
# list all the file with this extension with their full path
files <- list.files(flight_dir, pattern = ".csv.bz2", full.names = TRUE)
files
```


## Load data on SQLite

```{r}
# Try with 100 rows
max_rows_to_be_read <- 100
# When you are sure use them all, uncomment this line:
max_rows_to_be_read <- Inf

for (i in seq_along(files)) {
  append <- ifelse(i == 1, FALSE, TRUE)  # first time create the table, then append to the existing one
  overwrite <- ifelse(i != 1, FALSE, TRUE)
  this_file <- files[i]                  # work on the i-th file
  df <- read_delim(
    file = this_file, 
    delim = ",", 
    locale = locale(encoding = encoding),
    n_max = max_rows_to_be_read,
    show_col_types = FALSE
  )
  dbWriteTable(conn = con, name = "ontime", value = df ,append = append, overwrite = overwrite)
  cat(date() , this_file, "loaded", "\n")
  rm(df)
}
```

```{r}
dbGetQuery(
  con, "
  SELECT * FROM ontime LIMIT 100
  ")
```



```{r}
dbDisconnect(con)
```

## Use database with dplyr

### Lazy tables and lazy operations

```{r}
con_dplyr <- src_sqlite(path_db)
ontime <- tbl(con_dplyr, "ontime")
class(ontime)
dim(ontime)
```

Remark: a number `NA` of rows means that `ontime` is a lazy table.


```{r}
ontime_stat <- ontime %>% 
  group_by(Year, Month) %>%
  summarise(
    min = min(DepDelay, na.rm = TRUE), 
    avg = mean(DepDelay, na.rm = TRUE), 
    max = max(DepDelay, na.rm = TRUE),
    .groups = "drop"
  )
```


This operation has taken very little time. Because it is lazy. This means the real calculation happens when the result is requested:

```{r}
ontime_stat
```


Also the result of a dplyr operation is a lazy table:

```{r}
class(ontime_stat)
```


```{r}
show_query(ontime_stat)
```


## Create plots with collected data

### Collect a data frame

```{r}
ontime_dep_delay <- ontime %>%
  select(year = Year, dep_delay = DepDelay, arr_delay = ArrDelay) %>%
  # filter(dep_delay > 0) %>%
  collect()
```


This time the `collect()` function call immediately the computation and the collection of the result in the R process as a data frame.

```{r}
ontime_dep_delay
```


```{r}
class(ontime_dep_delay)
```



### Scatterplot with overplotting


```{r}
pl <- ontime_dep_delay %>% 
  head(n=5000) %>% 
  ggplot(aes(dep_delay, arr_delay)) + 
  geom_point(
    alpha=0.1
  )
```


Also ggplot is lazy, the plot is created just before being visualized:

```{r}
pl
```

### GGplot raster

```{r}
pl <- ggplot(ontime_dep_delay , aes(dep_delay, arr_delay)) + 
  geom_bin_2d(
    bins = 20
  ) 
  # facet_wrap(~year, ncol = 2)
pl
```

Note: in case you get a "Warning: Computation failed in `stat_binhex()`" run `install.packages("hexbin")`



## Create plots with pre-aggregated data by the database

### dbplot Raster quick function

```{r}
library(dbplot)

ontime_dep_delay_lazy <- ontime %>%
  select(
    year = Year,
    month = Month, 
    dep_delay = DepDelay, 
    arr_delay = ArrDelay, 
    origin = Origin,
    distance = Distance
  )

ontime_dep_delay_lazy %>%
  filter(!is.na(dep_delay)) %>% 
  dbplot_raster(dep_delay, arr_delay, resolution = 30)
```


### Raster with db_compute_raster()

```{r}
df <- ontime_dep_delay_lazy %>%
  db_compute_raster(x = dep_delay, y = arr_delay) %>% 
  collect() 

df %>%
  ggplot() +
  geom_bin_2d(
    mapping = aes(x = dep_delay, y = arr_delay, fill = `n()`),
    stat = "identity"
  )
```


### Raster with dplyr aggregation

```{r}
df <- ontime_dep_delay_lazy %>%
  group_by(
    dep_delay = !! db_bin(dep_delay, bins = 30),
    arr_delay = !! db_bin(arr_delay, bins = 30),
    year
  ) %>%
  summarise(avg_distance = mean(distance), .groups = "drop") %>%
  collect() 


df %>%
  ggplot() +
  geom_bin_2d(
    mapping = aes(x = dep_delay, y = arr_delay, fill = avg_distance),
    stat = "identity"
  ) +
  facet_wrap(~year, ncol = 2)
```


### Line plot

```{r}
ontime_dep_delay_lazy %>% 
  mutate(year_num = year + month/12) %>% 
  dbplot_line(year_num, mean(arr_delay)) +
  labs(title = "Arrival delays by year")
```


### Bar plot

```{r}
dbplot_bar(ontime_dep_delay_lazy, as.character(year), mean(dep_delay)) +
  labs(title = "Departure delays by year") +
  theme_minimal() +
  coord_flip()
```



### Histogram


```{r}
ontime_dep_delay_lazy <- ontime %>%
  select(
    year = Year, 
    dep_delay = DepDelay, 
    arr_delay = ArrDelay, 
    distance = Distance, 
    carrier = UniqueCarrier
  )
```


```{r}
ontime_dep_delay_lazy %>%
  group_by(x = !! db_bin(arr_delay)) %>%
  tally()
```

```{r}
df <- ontime_dep_delay_lazy %>%
  select(arr_delay) %>% 
  filter(!is.na(arr_delay)) %>%
  filter(arr_delay < 100 , arr_delay > -50) %>%
  group_by(x = !! db_bin(arr_delay, bins = 20)) %>%
  tally() %>%
  collect() 

df %>%
  ggplot() +
  geom_col(aes(x = x, y = n, fill = n))
```


