# Tidyr

```{r 06_setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(eval = T, message = F, warning = F)
```

## Overview
Date-time data can be frustrating to work with in R. R commands for date-times
are generally unintuitive and change depending on the type of date-time object
being used. Moreover, the methods we use with date-times must be robust to
time zones, leap days, daylight savings times, and other time related quirks, and
R lacks these capabilities in some situations. Lubridate makes it easier to do
the things R does with date-times and possible to do the things R does not.

Source: https://lubridate.tidyverse.org/

```{r message=F}
library(tidyverse)
library(lubridate)
library(ggplot2)
```


```{r}
# DATA
covid_raw <- read_delim("data/covid-ita-regions.csv", delim = ",")
covid_raw
```

```{r}
# DATA from the R format
covid_raw <- readRDS("data/covid-ita-regions.Rds")
covid_raw
```


## Parsing date and time
Usually dates and timestamps (date + day-time) are found in data in a wide variety of string formats, lubridate can convert very different date formats into date/timestamp format:

```{r}
# Extreme example from lubridate package help
x <- c(20100101120101,
"2009-01-02 12-01-02",
"2009.01.03 12:01:03",
"2009-1-4 12-1-4",
"2009-1, 5 12:1, 5",
"200901-08 1201-08",
"2009 arbitrary 1 non-decimal 6 chars 12 in between 1 !!! 6",
"OR collapsed formats: 20090107 120107 (as long as prefixed with zeros)",
"Automatic wday, Thu, detection, 10-01-10 10:01:10 and p format: AM",
"Created on 10-01-11 at 10:01:11 PM")
ymd_hms(x)
```


`ymd_hms` it is a function name which is acronym of Year, Month, Day, Hour, Minute, Second and it uses digits (from the input string) in this order to output a time object.

An example of conversion to a date format:

```{r}
# Parsing mechanism
x <- c("23/12/2013", "07/10/2014")
dmy(x)
```


If the information has different granularity the truncated param must be specified:


```{r}
# Handling truncation
x <- c("2011-12-31 12:59:59", "2010-01-01 12:11", "2010-01-01 12", "2010-01-01")
ymd_hms(x, truncated = 3) #up to 3 elements may be missing
```

```{r}
# Application on Covid data
covid_time <- covid_raw %>%
select(region = region_name, time) %>%
mutate(datetime = ymd_hms(time))
# Remark: If we don't specify the time zone, UTC is assumed
covid_time
```


## Working with time-zones


### Introduction

Consider timestamps in the Covid data set to be in the “CET” timezone (Central European Time). To interpret the time string in to the correct time zone the
param tz must be specified.


```{r}
# CONVERT TIME ZONES
covid_time %>%
mutate(datetime_tz = ymd_hms(time, tz = "CET"),
chicago_time = with_tz(datetime_tz, "America/Chicago"),
chicago_time_wrong = with_tz(datetime, "America/Chicago"))
# `chicago_time_wrong` is wrong, since it assumes we are in UTC timezone
```

### Accepted time-zones names
The list of accepted time-zones is returned by this function:

```{r}
# OlsonNames() # see help for more
head(OlsonNames(), n = 10)
```


### Fix time-zone of a timestamp

The function force_tz set a different timezone using the same time, it does not do the conversion, it only sets a different timezone.

```{r}
# FIX TIME-ZONES
# We can fix the initial mistake forcing the time-zone
covid_time <- covid_time %>%
mutate(datetime = force_tz(datetime, tzone = "Europe/Rome"))
```


## Setting and extracting info from a date and time objects


```{r}
covid_time %>%
  mutate(date = date(datetime),
         hour = hour(datetime),
         weekday = wday(date, label = T),
         month = month(date, label = T))
```

## Time span and arithmetic


### Time span

```{r}
# INTERVAL: time span between specific dates
covid_phase <- covid_time %>%
mutate(today = today(tzone = "CET"),
start_phase2 = ymd("2020-05-04", tz = "CET"),
phase2 = interval(start_phase2, today),
in_phase2 = datetime %within% phase2)
# Interval object have several helpers: int_start(), int_end(), int_shift() ...
# See interval() help for more

```


### Duration

```{r}
# DURATIONS: time duration in seconds
# Duration helper functions are named after the units of time, but start with a d
dminutes(3) # length of 2 minutes
class(dminutes(3))
```

### Period

```{r}

# PERIODS: time duration, adjusted to be more readable
# Period helper functions are named after the units of time, but plural
minutes(3)
class(minutes(3))
```

### Difference between Period and Duration

```{r}
# You can use duration and periods to do date arithmetic
ymd(20110101) + years(1) # More intuitive
ymd(20110101) + dyears(1) # More precise
```


```{r}
# Operation with dates
covid_phase <- covid_phase %>%
mutate(yesterday = today - days(1),
two_weeks_later = datetime + weeks(2),
length_phase2_d = phase2/days(1), # length in days
length_phase2_w = phase2/weeks(1) # length in weeks
)
# Months arithmetic is not so obvious.
# Try to be sure about what you want
# If adding or subtracting a month or a year creates an invalid date, lubridate will return an NA.
mar31 <- ymd(20190331)
mar31 + months(1)
# %m+% AND %m-% OPERATORS
# this avoid to outputs NAs
# If necessary, they roll back dates to the last day of the month
mar31 %m+% months(1)
mar31 %m-% months(1)
# If you want to add 31 days, then it's still different:
mar31 + days(31)
```

## Rounding dates

```{r}
covid_phase <- covid_phase %>%
mutate(date_month = floor_date(datetime, unit = "month"))
# Useful for summary stats
# grouping variable with lubridate -----------------------------------------
# In order to use a grouping variable of the month, you can create a string
# (concatenation of year and month), but in a more useful way you can use the
# computed last day of the month.
covid_time %>%
select(region, datetime) %>%
# mutate(mese = paste(year(datetime), month(datetime), sep = "-"))
mutate(date_month = ceiling_date(datetime, unit = "month") - days(1))
```

## Print time objects (convert to strings)


### Example print a nice label for a unique week of the year

```{r}
datetime <- ymd_hms("2011-12-31 12:59:59")
format(datetime, "%Y-W%U") # 2020-W01 Year-Week
# see help(strptime) for the %codes
# dates help(format.Date),
# date-times help(format.POSIXct)
```

### Using stamp
stamp help to create a format starting from an example string:

```{r}
# 1. Using an example string, create a template (a function)
# Tip: use a date with day > 12
example_date <- "Created Sunday, Jan 17, 1999 3:34"
sf <- stamp(example_date)
# 2. sf is a function to apply the template to dates
sf(ymd("2010-04-05"))
## [1] "Created Sunday, 04 05, 2010 00:00"
#> [1] "Created Monday, Apr 05, 2010 00:00"
```

