---
title: "Untitled"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Combine multiple data frames


### Joins

Joins are a way to merge table based on the correspondence on a key.

```{r}
library(tidyverse)
```

Let us a couple of example tables. Let us define a [foreign-key](https://en.wikipedia.org/wiki/Foreign_key) a column whose values have a correspondence in another table. This creates a relationship among two tables.

```{r}
# Main table, where
# 'id' id number
# 'lower' a foreign key
main_tbl <- tibble(
  id = c(1:2, 3, 4:6), #1:5,
  lower = c(letters[1:2], 'non-letter', letters[4:6])
)
main_tbl
```


And another that contains (usually all) the occurrences of the foreign-key:

```{r}
# The table of letters 
# 'lower' list all the lower case letters
# 'upper' an attribute to the letters, for the example case a upper case copy of the letter
letter_tbl <- tibble(
  lower = letters,
  upper = LETTERS
)
letter_tbl
```


```{r}
# full-join

main_tbl %>% 
  full_join(letter_tbl)
```


```{r}
# inner-join

main_tbl %>% 
  inner_join(letter_tbl)
```


```{r}
# left-join

main_tbl %>% 
  left_join(letter_tbl)
```


```{r}
# right-join

main_tbl %>% 
  right_join(letter_tbl)

# inverting the order a `right_join` return the same as a `left_join`
letter_tbl %>% 
  right_join(main_tbl)


```



### Split and Unite data frames

#### Bind rows

```{r}
mtcar_list <- split(mtcars, as.character(mtcars$cyl))
mtcars_cyl_6 <- mtcar_list[["6"]]
mtcars_cyl_8 <- mtcar_list[["8"]]
```



These two data frames have the same schema, and their rows can be concatenated:

```{r}
bind_rows(mtcars_cyl_6, mtcars_cyl_8)
```



#### Bind columns

```{r}
mtcars_mpg_cyl <- select(mtcars, mpg, cyl)
mtcars_disp_hp_drat <- select(mtcars, hp, drat, wt)
```

These two dataframes have the same number of rows, therefore their columns can be binded:

```{r}
bind_cols(mtcars_disp_hp_drat, mtcars_mpg_cyl)
```

