---
title: "03 Describe a categorical variable"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggplot2)
```

## Dataset

```{r}
class_student <- tribble(
  ~Name, ~Gender, ~Grade, ~Age,
  "Jude",     "M",    "A",  27L,
  "Philip",     "M",    "B",  24L,
  "Carl",     "M",    "A",  41L,
  "Anne",     "F",    "A",  32L,
  "Rose",     "F",    "B",  26L,
  "Peter",     "M",    "C",  30L,
  "Mary",     "F",    "A",  45L,
  "Michaela",     "F",    "B",  26L,
)
```


## Frequency table (categorical variable)

The frequency table of a *categorical* variable is a tables with the following aggregations relating to the specified variable:

- *Absolute frequency*: the number of times a specific level value occurs.

- *relative frequency*: the absolute frequency normalized by the total number of records. 

- *percentage*: the relative frequency multiplied by 100.

- *cumulative absolute frequency*: the total of the *absolute* frequencies of all records at or below a certain point in an ordered list of events (Advice: use the function `cumsum` in R).

- *cumulative relative frequency*: the total of the *relative* frequencies of all records at or below a certain point in an ordered list of events (Advice: use the function `cumsum` in R).

- *cumulative percentage*  the cumulative frequency multiplied by 100.


Remark: cumulative aggregation are meaningful if the variables level can be ordered.


**Exercise**: write the frequency table of the `class_student$Grade` variable:

```{r}
tot <- nrow(class_student)

class_student %>%
  group_by(
    Grade
  ) %>% 
  summarise(
    abs_freq = n(),
    rel_freq = abs_freq / tot,
    perc = rel_freq * 100
  ) %>% 
  mutate(
    cum_abs_freq = cumsum(abs_freq),
    cum_rel_freq = cumsum(rel_freq),
    cum_perc = cumsum(perc)
  )
```

## Bar plot 

Bar plot is a graphical visualization that has the levels of a categorical variable at the x-axis and a per-level aggregation on the y-axis (bin high).

The default aggregation for the `geom_bar` function is the count:

```{r}
ggplot(class_student) +
  aes(x = Grade) +
  geom_bar(fill = "darkgreen")
```

In case you need a different aggregation you can calculate it yourself:

```{r}
class_student %>% 
  group_by(Grade) %>% 
  summarise(
    rel_freq = n()/tot
  ) %>% 
  ggplot() +
  aes(x = Grade, y = rel_freq) +
  geom_bar(stat = "identity", fill = "darkgreen") +
  scale_y_continuous(labels = scales::label_percent()) +
  ylab(label = "Percentage (%)")
```


## Bivariate

Two way contingency table:

```{r}
table(class_student$Gender, class_student$Grade, dnn = c("Gender", "Grade"))
```


Stacked barplot:

```{r}
ggplot(class_student) +
  aes(x = Grade, fill = Gender) +
  geom_bar(position = 'stack')
```

Dodge barplot:

```{r}
ggplot(class_student) +
  aes(x = Grade, fill = Gender) +
  geom_bar(position = 'dodge')
```


