---
title: "Pivoting"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Tidy data (definition)

Tidy data is a standard way of mapping the meaning of a dataset to its structure. A dataset is messy or tidy depending on how rows, columns and tables are matched up with observations, variables and types. In tidy data:

    Every column is a variable.

    Every row is an observation.

    Every cell is a single value.

See also the [tidyr vignette](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html).


## Pivot Longer

```{r}
library(tidyverse) 
# library(tidyr)
library(dslabs)
path <- system.file("extdata", package="dslabs")
filename <- file.path(path, "fertility-two-countries-example.csv")
wide_data <- read_csv(filename)
wide_data
```


```{r}
new_tidy_data <- wide_data %>% 
  pivot_longer(`1960`:`2015`, names_to = "year", values_to = "fertility")
new_tidy_data
```


## Gapminder

```{r}
data("gapminder")
tidy_data <- gapminder %>% 
  filter(country %in% c("South Korea", "Germany") & !is.na(fertility)) %>%
  select(country, year, fertility)
tidy_data
```

```{r}
new_wide_data <- tidy_data %>% 
  pivot_wider(names_from = year, values_from = fertility)

new_wide_data
```


## Separate 

```{r}
path <- system.file("extdata", package = "dslabs")

filename <- "life-expectancy-and-fertility-two-countries-example.csv"
filename <-  file.path(path, filename)

raw_dat <- read_csv(filename)
select(raw_dat, 1:5)
```


```{r}
dat <- raw_dat %>% 
  pivot_longer(-country)
dat
```


```{r}
dat %>% separate(name, c("year", "name"), "_", extra = "merge")
```

```{r}
# Finally the tidy dataframe
dat %>% 
  separate(name, c("year", "name"), extra = "merge") %>%
  pivot_wider()
```

## Unite

```{r}
var_names <- c("year", "first_variable_name", "second_variable_name")
df <- dat %>% 
  separate(name, var_names, fill = "right")
df
```


```{r}
df %>%
  unite(col = name, first_variable_name, second_variable_name, sep = "_") %>%
  pivot_wider() %>%
  rename(fertility = fertility_NA)
```

